Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-06-17 manuscript/manuscript.Rmd
updates ms discussion

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
mainfont: times
geometry: margin=1in

@@

Microscale models outperform mesoscale models when forecasting climate change impacts on plant populations
========================================================================================

@@

Discussion
----------
Demographic models are costly to parameterize, yet remain the most widely-used tools for studying plant population dynamics (e.g., @Salguero-Gomez2015). New density-structured approaches sacrifice some precision at the local scale for better spatial coverage that is virtually unattainable, due to time and funding constraints, using traditional individual-based data [@Freckleton2011; @Queenborough2011]. These approaches are appealing because traditional population models suffer from constrained parameter space -- meaning forecasts quickly extend beyond the parameter space used to fit the model. Density-structured approaches rely on easy-to-collect census data (e.g., percent cover in $1\text{m}^2$ quadrats) so sampling populations across large spatial extents would be possible, thus covering more potential parameter space.

Estimating parameter variability in space and time is critical when using population models to forecast responses to directional changes in exogoneous drivers like climate [@Clark2011; @Vanderwel2013; @Wolkovich2014; @Petchey2015]. Compared to demographic models, density-structured modeling approaches may provide an efficient route toward estimating such uncertainty. However, population models based on aggregated inidividual data have yet to be tested with climate drivers. Since it is individuals that respond to weather, rather than populations, models based on a aggregated data may produce incorrect forecasts [@Clark2012]. Thus, we sought to compare climate sensitivities of individual and population based models of four grassland species populations.

@@

>>>>

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in

@@

Do we need detailed demographic data to forecast the impacts of climate change on populations?
========================================================================================

@@

Discussion
----------
Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001]. To do so requires sophisticated modeling approaches that fully account for uncertainty and variability in the ecological process and associated parameters. This requires large amounts of data collected over large spatio-temporal extents. State-of-the-science modeling techniques cannot overcome data limitations. Such is the case for many population models.

As a potential remedy to the "data dearth" problem, @Queenborough2011 and @Freckleton2011, building on work by @Taylor2004, advocate a "density-structured" modeling approach. Such models do not require individual level demographic data. Indeed, @Queenborough2011 provide empirical evidence that density-structured models adequately described population dynamics. The results from density-structured models are not as precise as those from traditional population models, but the loss in precision is traded off with a gain in data. The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks. This is far more data from a far greater spatial extent than possible if measuring individual plant demography (in a world of limited time and money, at least). The appeal of density-structured approaches is clear.

However, at their core, density-structured models rely on individual level data aggregated to a population level metric (e.g., density classes or percent cover). This creates a potential problem if such models are to be used in a climate change context because inidividuals respond to climate, not populations [@Clark2012]. Are models based on population level metrics as sensitive to climate as models based on individual level metrics? Do these two types of models produce consistent forecasts? Do we need detailed demographic data to forecast the impacts of climate change? These are the questions we sought to answer here.

Using individual and aggregated forms of the same dataset, we were able to directly compare a traditional demographic modeling approach to a population model based on aggregated data. Our quad-based model (QBM) is based on percent cover data and so is in the spirit of density-structured models. In terms of each model's forecasting ability, the IPM outperformed the QBM (Table 1). This is unsurprising since we expected to lose some precision at the aggregated level. However, the underwhelming performance of the QBM could call in the question forecasts that differ from the IPM.

Indeed, when we perturbed climate factors the QBM made forecasts completely contradictory to those of the IPM for three of our four study species (Fig. 2a-d). In a perfect world, the QBM would have made forecasts of at least the same direction as the IPM. If that had been the case we could conclude that aggregate level models could prove useful for forecasting climate change impacts on populations. Unfortunately, this was not the case.

Given the superior ability of the IPM to predict out of sample observations (Table 1), we have no choice but to conclude it is the superior model. Following that logic, we can only assume that, at least contingent on the data in hand, the IPM is producing the correct forecasts to climate perturbations. The QBM failed to match IPM forecasts, implying that detailed demographic data may be necessary to accurately detect climate signals that are utlimately important at the population level. This result further confirms related work on the importance of individual variability on population level responses to exogenous drivers (CITATIONS).

This result is not as dire as it seems when we consider model uncerainty. Yes, the QBM failed to match IPM forecasts under perturbed climate, but the uncertainties around those forecasts were so great (Fig. 2e-h) that we cannot say that the two model's forecasts were different with any level of confidence. This offers hope that with "enough" population level data one will be able to reduce parameter uncertainty and thus model forecasts. This remains an open question.

@@

Conclusions
-----------

This work is not a critique of density-structured population models. Instead, we consider our results a cautionary tale. In some cases and for certain species, population models based on aggregated data may prove useful and unbiased. But, as our work shows, that is not a foregone conclusion when considering the impact of exogenous drivers like climate on population states. Thus, forecasts from aggregate level models should be viewed with caution and should never be unaccompianed by uncertainty. As we have shown, uncertainty around forecasts was large for both model types. Which leads us to our most pessimistic conclusion: even with 14 years of detailed demographic data and sophisticated modeling techniques we failed to produce forecasts with any level of acceptable uncertainty. In our view, uncertainty of climate change related forecasts can be reduced by (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics). Do we need detailed demographic data to forecast the impacts of climate change on populations? Probably.

### 2015-06-17 manuscript/manuscript.tex
updates ms discussion

\documentclass[12pt,]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
    \setmainfont{times}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
% Redefine \includegraphics so that, unless explicit options are
% given, the image width will not exceed the width of the page.
% Images get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\ScaleIfNeeded{%
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother
\let\Oldincludegraphics\includegraphics
{%
 \catcode`\@=11\relax%
 \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
}%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\section{Microscale models outperform mesoscale models when forecasting
climate change impacts on plant
populations}\label{microscale-models-outperform-mesoscale-models-when-forecasting-climate-change-impacts-on-plant-populations}

@@

Demographic models are costly to parameterize, yet remain the most
widely-used tools for studying plant population dynamics (e.g.,
Salguero-Gómez et al. (2015)). New density-structured approaches
sacrifice some precision at the local scale for better spatial coverage
that is virtually unattainable, due to time and funding constraints,
using traditional individual-based data (Freckleton et al. 2011,
Queenborough et al. 2011). These approaches are appealing because
traditional population models suffer from constrained parameter space --
meaning forecasts quickly extend beyond the parameter space used to fit
the model. Density-structured approaches rely on easy-to-collect census
data (e.g., percent cover in $1\text{m}^2$ quadrats) so sampling
populations across large spatial extents would be possible, thus
covering more potential parameter space.

Estimating parameter variability in space and time is critical when
using population models to forecast responses to directional changes in
exogoneous drivers like climate ({\textbf{???}}, {\textbf{???}},
{\textbf{???}}, {\textbf{???}}). Compared to demographic models,
density-structured modeling approaches may provide an efficient route
toward estimating such uncertainty. However, population models based on
aggregated inidividual data have yet to be tested with climate drivers.
Since it is individuals that respond to weather, rather than
populations, models based on a aggregated data may produce incorrect
forecasts (Clark et al. 2012). Thus, we sought to compare climate
sensitivities of individual and population based models of four
grassland species populations.

@@

@@

>>>>

\documentclass[12pt,]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
% Redefine \includegraphics so that, unless explicit options are
% given, the image width will not exceed the width of the page.
% Images get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\ScaleIfNeeded{%
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother
\let\Oldincludegraphics\includegraphics
{%
 \catcode`\@=11\relax%
 \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
}%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\section{Do we need detailed demographic data to forecast the impacts of
climate change on
populations?}\label{do-we-need-detailed-demographic-data-to-forecast-the-impacts-of-climate-change-on-populations}

@@

Perhaps the greatest challenge for ecology in the 21st century is to
forecast the impacts of environmental change (Clark et al. 2001). To do
so requires sophisticated modeling approaches that fully account for
uncertainty and variability in the ecological process and associated
parameters. This requires large amounts of data collected over large
spatio-temporal extents. State-of-the-science modeling techniques cannot
overcome data limitations. Such is the case for many population models.

As a potential remedy to the ``data dearth'' problem, Queenborough et
al. (2011) and Freckleton et al. (2011), building on work by Taylor and
Hastings (2004), advocate a ``density-structured'' modeling approach.
Such models do not require individual level demographic data. Indeed,
Queenborough et al. (2011) provide empirical evidence that
density-structured models adequately described population dynamics. The
results from density-structured models are not as precise as those from
traditional population models, but the loss in precision is traded off
with a gain in data. The study by Queenborough et al. (2011) included
data from 500 fields (4 hectares each) in 49 farms, all collected by two
people in 6 weeks. This is far more data from a far greater spatial
extent than possible if measuring individual plant demography (in a
world of limited time and money, at least). The appeal of
density-structured approaches is clear.

However, at their core, density-structured models rely on individual
level data aggregated to a population level metric (e.g., density
classes or percent cover). This creates a potential problem if such
models are to be used in a climate change context because inidividuals
respond to climate, not populations (Clark et al. 2012). Are models
based on population level metrics as sensitive to climate as models
based on individual level metrics? Do these two types of models produce
consistent forecasts? Do we need detailed demographic data to forecast
the impacts of climate change? These are the questions we sought to
answer here.

Using individual and aggregated forms of the same dataset, we were able
to directly compare a traditional demographic modeling approach to a
population model based on aggregated data. Our quad-based model (QBM) is
based on percent cover data and so is in the spirit of
density-structured models. In terms of each model's forecasting ability,
the IPM outperformed the QBM (Table 1). This is unsurprising since we
expected to lose some precision at the aggregated level. However, the
underwhelming performance of the QBM could call in the question
forecasts that differ from the IPM.

Indeed, when we perturbed climate factors the QBM made forecasts
completely contradictory to those of the IPM for three of our four study
species (Fig. 2a-d). In a perfect world, the QBM would have made
forecasts of at least the same direction as the IPM. If that had been
the case we could conclude that aggregate level models could prove
useful for forecasting climate change impacts on populations.
Unfortunately, this was not the case.

Given the superior ability of the IPM to predict out of sample
observations (Table 1), we have no choice but to conclude it is the
superior model. Following that logic, we can only assume that, at least
contingent on the data in hand, the IPM is producing the correct
forecasts to climate perturbations. The QBM failed to match IPM
forecasts, implying that detailed demographic data may be necessary to
accurately detect climate signals that are utlimately important at the
population level. This result further confirms related work on the
importance of individual variability on population level responses to
exogenous drivers (CITATIONS).

This result is not as dire as it seems when we consider model
uncerainty. Yes, the QBM failed to match IPM forecasts under perturbed
climate, but the uncertainties around those forecasts were so great
(Fig. 2e-h) that we cannot say that the two model's forecasts were
different with any level of confidence. This offers hope that with
``enough'' population level data one will be able to reduce parameter
uncertainty and thus model forecasts. This remains an open question.

@@

\subsection{Conclusions}\label{conclusions}

This work is not a critique of density-structured population models.
Instead, we consider our results a cautionary tale. In some cases and
for certain species, population models based on aggregated data may
prove useful and unbiased. But, as our work shows, that is not a
foregone conclusion when considering the impact of exogenous drivers
like climate on population states. Thus, forecasts from aggregate level
models should be viewed with caution and should never be unaccompianed
by uncertainty. As we have shown, uncertainty around forecasts was large
for both model types. Which leads us to our most pessimistic conclusion:
even with 14 years of detailed demographic data and sophisticated
modeling techniques we failed to produce forecasts with any level of
acceptable uncertainty. In our view, uncertainty of climate change
related forecasts can be reduced by (1) longer time series and (2)
climate covariates that match the scale of inference (e.g., plot rather
than landscape level climate/weather metrics). Do we need detailed
demographic data to forecast the impacts of climate change on
populations? Probably.

@@

Clark, J. S., S. R. Carpenter, M. Barber, S. Collins, A. Dobson, J. A.
Foley, D. M. Lodge, M. Pascual, R. Pielke, W. Pizer, C. Pringle, W. V.
Reid, K. A. Rose, O. Sala, W. H. Schlesinger, D. H. Wall, and D. Wear.
2001. Ecological forecasts: an emerging imperative. Science (New York,
N.Y.) 293:657--660.
